---
title: "Intro to R, RStudio, and Tidyverse"
author: "Chris Castañeda-Barajas"
output: html_notebook
---

# Goals:
- Know enough R to be dangerous!
- Understand why you'd want to use R, and what it is exactly.

- Cover R Bacis:
    - variables, vectors & data frames
    - Functions: What is a function? built in functions
    - Loadings a file
    - Preform basic operations: 
      sum, mean, min, max, sd, summary
    - Know where to look to do more coplex opperations
    - Working with data frames

- The Tidyverse!
    - ggplot2, dplyr, tidyr, readr, purrr, tibble, stringr, forcats

# Installing Libraries


```{r}
# install.packages(c("tidyverse", "reshape2"))
```

```{r}
library(tidyverse)  
library(reshape2)
```


# Variables

```{r}
# Signle Values
x <- 42  
y <- "forty two"
z <- TRUE
```

```{r}
# Arrays / Vectors
a <- c(1, 2, 3, 4, 5)
b <- c(x, y, z)
c <- c("Seattle", "San Francisco", "Seattle", "Portland", "Tacoma")
```


```{r}
a[1]
b[1]
c[1]
```

```{r}
a[1:3]
```

```{r}
a[2:4]
```

```{r}
c[c(1,3,5)]
```


# Fun with functions!

```{r}
# Sum
sum(a)
```

```{r}
# Length
length(a)
```

```{r}
# Calculate Average
sum(a) / length(a)
```



```{r}
# Average
mean(a)
```

```{r}
# Median
median(a)
```

```{r}
# Standard Deviation
sd(a)
```

```{r}
# Min value
min(a)

# Max value
max(a)
```

```{r}
# Summary function
summary(a)
```




```{r}
# Writing your own functions
add_two <- function(x) {
  x + 2
}
```

```{r}
add_two(x)
```

```{r}
add_two(a)
```

```{r}
mean(add_two(a))
```

```{r}
summary(add_two(a))
```



## Using Pipes `%>%`

```{r}
a %>% add_two
```

```{r}
a %>% add_two %>% mean
```


## Other Useful Functions
```{r}
## Unique and Length
c 
c %>% length
c %>% unique
c %>% unique %>% length
```

```{r}
## Determine Data Type
c %>% class
c %>% is
```


```{r}
## Paste
c %>% paste
```



# Working with Data Frames

```{r}
## Make a data frame from scratch
df <- data_frame(
  col.1 = c(1, 2, 3, 4, 5, 6, 7, 8, 9),
  col.2 = c("a", "b", "c", "d", "e", "f", "g", "h", "i"),
  col.3 = c(TRUE, TRUE, TRUE, FALSE, TRUE, FALSE, FALSE, TRUE, NA)
)
```

```{r}
## Load dataframe from a spreadsheet
demo.donors <- read_csv("data/demo_donors.csv")
demo.transactions <- read_csv("data/demo_transactions.csv")
```

```{r}
## View the data frame
df
```

```{r}
View(df)
```

```{r}
# Dimentions of a data frame
df %>% dim
df %>% nrow
df %>% ncol
```

```{r}
# Column Names
df %>% names
```

```{r}
# Renaming columns
names(df) <- c("Numbers", "Letters", "Booleans")
```

```{r}
df %>% names
```

```{r}
df
```


```{r}
# View / Use only certain rows
df[1,]
```

```{r}
df[1:8,]
```

```{r}
df[3,2]
```

```{r}
df[,1]
```

```{r}
df$Letters
```

```{r}
df$Booleans
```

```{r}
# Finding N/A Values
df$Booleans %>% is.na
df$Booleans %>% is.na %>% sum
```


# Working with the demo data

## Filtering

```{r}
demo.donors %>%
  filter(CITY == "SEATTLE")
```

```{r}
demo.donors %>%
  filter(CITY %in% c("FEDERAL WAY", "AUBURN"))
```

## Pivot Tables
```{r}
demo.donors %>%
  dcast(CITY ~ .)
```

```{r}
demo.donors %>%
  dcast(CITY ~ .) %>%
  arrange(desc(.))
```




## Joins

```{r}
# Take a look at demo.transactions data frame
demo.transactions
```


```{r}
demo.transactions %>% 
  left_join(demo.donors, by="CONS_ID")
```


### Question: How much money was raised by each city? 


```{r}
demo.transactions %>% 
  left_join(demo.donors, by="CONS_ID") %>%
  dcast(CITY ~ ., value.var = c("GIFT_AMOUNT"), fun.aggregate = sum) 
  # set_colnames(c("CITY", "GIFT_AMOUNT")) %>%
  # arrange(desc(GIFT_AMOUNT))
```


# Plots and data viz!

```{r}
demo.transactions %>% 
  left_join(demo.donors, by="CONS_ID") %>%
  dcast(CITY ~ ., value.var = c("GIFT_AMOUNT"), fun.aggregate = sum) %>%
  set_colnames(c("CITY", "GIFT_AMOUNT")) %>%
  arrange(desc(GIFT_AMOUNT)) %>%
  # filter(CITY %in% c("FEDERAL WAY", "AUBURN")) %>%
  ggplot(aes(x=CITY, y=GIFT_AMOUNT)) + 
    geom_col()
```

## Histograms

```{r}
demo.transactions %>% 
  ggplot(aes(GIFT_AMOUNT)) +
    geom_histogram() #+ 
      # facet_wrap(~ CAMPAIGN_NAME)
```


